{\huge Changes and trends in budburst and leaf flush across Europe and North America} \\A meta-analysis of local adaptation in spring phenology studies

### 2023-04-04 Docs/manuscript.Rnw
edited results

Overall, spring events were not related to provenance latitude (\Sexpr{format(mean(northamerican_springpost_fitC_spring_lat),digits = 1)} [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)}-\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]) or MAT (\Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)}-\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) in North America. \todo{add figure here} Spring events advanced slightly with provenance latitude (\Sexpr{format(mean(european_springpost_fitC_spring_lat),digits = 1)} [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)}-\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]) and MAT (\Sexpr{format(mean(european_springpost_fitC_spring_mat),digits = 1)} [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)}-\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]) in Europe, meaning spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes)(Fig. \ref{figure:springfall_latmat} \& Supplement Table. \ref{supplement-table:model_sf_lat}), though uncertainty intervals strongly overlapped zero. Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the absolute value of the distance and the spherical distance difference between provenance and garden latitude).

@@

In contrast, fall events (budset, leaf senescence, leaf abcission) advanced strongly with provenance latitude and MAT, meaning fall events are earlier where provenance MAT is lower. This relationship, however, was observed mostly in North America where fall events advance \Sexpr{format(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2)} (\Sexpr{format(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2)}-\Sexpr{format(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2)}) days per degree northward, or \Sexpr{format(mean(northamericanpost_fitC_fall_mat),digits = 2)} days (\Sexpr{format(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2)}-\Sexpr{format(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2)}) per degree decline in MAT$^{\circ}$C (Fig. \ref{figure:continent}). In Europe, such relationship is weak: advance \Sexpr{format(abs(mean(europeanpost_fitC_fall_lat)),digits = 1)} (\Sexpr{format(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 1)}-\Sexpr{format(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2)}) days per degree northward, or \Sexpr{format(mean(europeanpost_fitC_fall_mat),digits = 1)} days (\Sexpr{format(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2)}-\Sexpr{format(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2)}) per degree decline in MAT $^{\circ}$C(Fig. \ref{figure:continent}).

@@

Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms. However, effects of MAT on spring events weakly diverge: spring events get earlier as MAT increases in angiosperms and delay as MAT increases in gymnosperms, except for \emph{Pseudotsuga menziesii}. Fall events delay in warmer locations for both species types, but slightly more so for gymnosperms (\Sexpr{format(mean(deciduouspost_fitC_fall_mat),digits = 2)} days VS. \Sexpr{format(mean(evergreenspost_fitC_fall_mat),digits = 2)} days)(Fig. \ref{figure:spp_type}).

>>>>

Overall, spring events were not related to provenance latitude (\Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]) or MAT (\Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) in North America. \todo{add figure here} Spring events advanced slightly with provenance latitude (\Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]) and MAT (\Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]) in Europe, meaning spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes)(Fig. \ref{figure:springfall_latmat} \& Supplement Table. \ref{supplement-table:model_sf_lat}), though uncertainty intervals strongly overlapped zero. Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the absolute value of the distance and the spherical distance difference between provenance and garden latitude).

@@

In contrast, fall events (budset, leaf senescence, leaf abcission) advanced strongly with provenance latitude and MAT, meaning fall events are earlier where provenance MAT is lower. This relationship, however, was observed mostly in North America where fall events advance \Sexpr{format(round(abs(mean(northamericanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(northamericanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(northamericanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT$^{\circ}$C (Fig. \ref{figure:continent}). In Europe, such relationship is weak: advance \Sexpr{format(round(abs(mean(europeanpost_fitC_fall_lat)),digits = 2),nsmall=2)} (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_lat, 0.05)), digits = 2),nsmall=2)}) days per degree northward, or \Sexpr{format(round(mean(europeanpost_fitC_fall_mat),digits = 2),nsmall=2)} days (\Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(europeanpost_fitC_fall_mat, 0.95)), digits = 2),nsmall=2)}) per degree decline in MAT $^{\circ}$C(Fig. \ref{figure:continent}).

@@

Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms. However, effects of MAT on spring events weakly diverge: spring events get earlier as MAT increases in angiosperms and delay as MAT increases in gymnosperms, except for \emph{Pseudotsuga menziesii}. Fall events delay in warmer locations for both species types, but slightly more so for gymnosperms (\Sexpr{format(round(mean(deciduouspost_fitC_fall_mat),digits = 2),nsmall=2)} days VS. \Sexpr{format(round(mean(evergreenspost_fitC_fall_mat),digits = 2),nsmall=2)} days)(Fig. \ref{figure:spp_type}).

### 2023-04-04 Docs/manuscript.tex
edited results

Overall, spring events were not related to provenance latitude (0.1 [-0.06-0.26]) or MAT (-0.09 [-0.36-0.18]) in North America. \todo{add figure here} Spring events advanced slightly with provenance latitude (0.1 [-0.23-0.42]) and MAT (-0.2 [-0.55-0.23]) in Europe, meaning spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes)(Fig. \ref{figure:springfall_latmat} \& Supplement Table. \ref{supplement-table:model_sf_lat}), though uncertainty intervals strongly overlapped zero. Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the absolute value of the distance and the spherical distance difference between provenance and garden latitude).

@@

In contrast, fall events (budset, leaf senescence, leaf abcission) advanced strongly with provenance latitude and MAT, meaning fall events are earlier where provenance MAT is lower. This relationship, however, was observed mostly in North America where fall events advance 4.2 (3.9-4.5) days per degree northward, or 6.4 days (5.8-7) per degree decline in MAT$^{\circ}$C (Fig. \ref{figure:continent}). In Europe, such relationship is weak: advance 0.5 (0.2-1.2) days per degree northward, or 0.7 days (1-2.4) per degree decline in MAT $^{\circ}$C(Fig. \ref{figure:continent}).

@@

Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms. However, effects of MAT on spring events weakly diverge: spring events get earlier as MAT increases in angiosperms and delay as MAT increases in gymnosperms, except for \emph{Pseudotsuga menziesii}. Fall events delay in warmer locations for both species types, but slightly more so for gymnosperms (3.7 days VS. 6.2 days)(Fig. \ref{figure:spp_type}).

>>>>

Overall, spring events were not related to provenance latitude (0.10 days/degree [-0.06 - 0.26]) or MAT (-0.09 days/degree [-0.36 - 0.18]) in North America. \todo{add figure here} Spring events advanced slightly with provenance latitude (0.10 days/degree [-0.23 - 0.42]) and MAT (-0.16 days/degree [-0.55 - 0.23]) in Europe, meaning spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes)(Fig. \ref{figure:springfall_latmat} \& Supplement Table. \ref{supplement-table:model_sf_lat}), though uncertainty intervals strongly overlapped zero. Results were similar using other distance metrics in lieu of latitude (see Supplement Fig.\ref{supplement-figure:lat_distance} for results using the absolute value of the distance and the spherical distance difference between provenance and garden latitude).

@@

In contrast, fall events (budset, leaf senescence, leaf abcission) advanced strongly with provenance latitude and MAT, meaning fall events are earlier where provenance MAT is lower. This relationship, however, was observed mostly in North America where fall events advance 4.24 (3.95 - 4.53) days per degree northward, or 6.41 days (5.78 - 7.04) per degree decline in MAT$^{\circ}$C (Fig. \ref{figure:continent}). In Europe, such relationship is weak: advance 0.47 (0.21 - 1.17) days per degree northward, or 0.70 days (1.04 - 2.42) per degree decline in MAT $^{\circ}$C(Fig. \ref{figure:continent}).

@@

Effects of provenance latitude on both spring and fall events were similar across angiosperms and gymnosperms. However, effects of MAT on spring events weakly diverge: spring events get earlier as MAT increases in angiosperms and delay as MAT increases in gymnosperms, except for \emph{Pseudotsuga menziesii}. Fall events delay in warmer locations for both species types, but slightly more so for gymnosperms (3.69 days VS. 6.23 days)(Fig. \ref{figure:spp_type}).
